Staged Abstract Interpreters

### 2019-02-19 paper/10_intro.tex
intro

\todo{example of interpreter -> compiler, mix}

@@

Now we have seen two orthogonal transformations of interpreters: by
speclization, we can refactor interpreters to compilers; by abstraction, we can
refactor interpreters to static analyzers. As an intellectual quest, it is
natural to ask --- starting from an interpreter, can we derive sound abstract
interpreters but it works like a compiler and generates fast analysis code?

@@

\begin{figure}[h]
  \begin{tikzpicture}
  \matrix (m) [matrix of math nodes,row sep=3em,column sep=4em,minimum width=2em]
  {
    \begin{smallmatrix} \text{unstaged} \\ \text{abstract} \\ \text{interpreters} \end{smallmatrix} & \begin{smallmatrix} \text{staged} \\ \text{abstract} \\ \text{interpreters} \end{smallmatrix} \\
      \begin{smallmatrix} \text{unstaged} \\ \text{concrete} \\ \text{interpreters} \end{smallmatrix} & \begin{smallmatrix} \text{staged} \\ \text{concrete} \\ \text{interpreters} \end{smallmatrix} \\ };
  \path[-stealth]
    (m-1-1) edge node [above, font=\scriptsize] {$\updownarrows$} (m-1-2)
    (m-2-1) edge node [left, font=\scriptsize] {$\alpha$} (m-1-1)
    (m-2-1) edge node [below, font=\scriptsize] {$\updownarrows$} (m-2-2)
    (m-2-2) edge node [right, font=\scriptsize] {$\alpha$} (m-1-2);
  \end{tikzpicture}
  \caption{The confluence of staging and approximation.}
  \label{confluence}
\end{figure}

Besides the intellectual merit of the confluence diagram, this paper contributes
to a practical problem in static analysis: can we speed up a static analyzer
without any intrusive changes and soundness compromises?
\todo{specializing analysis wrt program is not new; first staging; we argue that
staging is less intrusive compared with AC (ie, closure generation), in
fact,LMS-based staging only requires type annotations }

\todo{Move some into related work}
Multi-stage programming has been widely used to improve the performance in many
domains, such as optimizing compilers or domain-specific languages
\cite{DBLP:conf/pldi/RompfSBLCO14, DBLP:conf/snapl/RompfBLSJAOSKDK15,
DBLP:journals/tecs/SujeethBLRCOO14, DBLP:conf/gpce/SujeethGBLROO13,
DBLP:journals/jfp/CaretteKS09}, numerical computation \cite{PGL-038,
DBLP:conf/pepm/AktemurKKS13}, generic programming
\cite{DBLP:journals/pacmpl/Yallop17, Ofenbeck:2017:SGP:3136040.3136060}, data
processing \cite{DBLP:conf/oopsla/JonnalageddaCSRO14,
DBLP:conf/popl/KiselyovBPS17}, query compilation in databases
\cite{DBLP:conf/osdi/EssertelTDBOR18, DBLP:conf/sigmod/TahboubER18}, etc.
Likewise, specializing static analyses by partial evaluation emerged in late 90s
\cite{damian1999partial, amtoft1999partial, Boucher:1996:ACN:647473.727587,
ashley:practical}, and indeed it is able to effectively remove the interpretive
overhead of repeatedly traversing the abstract syntax tree. However, these
previous works focused mostly on one particular analysis, or required to
completely rewrite the analyzer. Hence, it is worth to investigate the idea from
a modern perspective based on generative programming, especially for a general
abstract interpreter that models direct-style $\lambda$-calculus, imperative
features such as mutation, and different abstract domains. One technical
challenge is the binding-time engineering when non-determinism, fixed-point
iterations, different abstract domains and staging are introduced at the same
time. In this paper, we present an end-to-end design and implementation of
staging an abstract interpreter; that means not only the interpreter that
traverses the abstract syntax tree, but also the data structure of abstract
domains, abstract environment and heap, and the fixed-point iteration are
staged.

@@

\iffalse
When staging a concrete interpreter, the programmers need to distinguish static
and dynamic values --- the given program to be executed by the interpreter is
classified as static because it is known at compile-time, and the inputs to that
program are dynamic. However, when staging an abstract interpreter, this
distinction does not exist anymore. Because the abstract interpreter
instantiates all the inputs as some form of abstract values, which are usually
top elements in their abstract domains and are also statically known. Then what
is the point of staging if there is no such distinction? A surprising by-product
of thinking about this question is to realize that we can apply the staged
abstract interpreter on \textit{open} programs, and the free variables
representing other parts of the program (e.g., libraries) are dynamic inputs,
therefore we obtain a modular analysis through staging, mechanically.
\note{TR: I don't understand this. The program structure is static, the abstract values
are still dynamic, no? They change in every iteration of the fixpoint algorithm}
\note{TR: why is it a big deal. On Closed programs we obtain a constant factor,
but modular analysis has different asymptotics}
\fi

>>>>

Now consider that we have an interpreter @eval@ for some language as following:

\begin{lstlisting}
def eval(e: Expr)(arg: Input): Value = ...
\end{lstlisting}

where $e$ is the program of type @Expr@ and @arg@ is the input to program $e$,
and @eval@ produces a value.
Semantically, the interpreter satisfies $ \texttt{eval}(e)(arg) = [\![ e ]\!] arg$ for all
programs and inputs. Given a program $e_0$, by applying the first Futamura
projection, we may obtain a specialized interpreter
$\texttt{eval}_{\texttt{e0}}$ with respect to $e_0$, and by definition of the interpreter, 
$\texttt{eval}_{\texttt{e0}}(arg) = [\![ e_0 ]\!] arg $. If we generalize the
argument @arg@ to an environment (as well as a companion heap object, if for the
interest of using assginments) that contains values mapped from free variables
in $e$, the interpreter is a standard environment/store-passing interpreter,
where the specialization through MSP is also applicable.

@@

Now we have seen two orthogonal transformations of interpreters: by
speclization, we can refactor interpreters to compilers; by abstraction, we can
refactor interpreters to static analyzers. As an intellectual quest, it is
natural to ask --- starting from an interpreter, \textbf{can we derive sound abstract
interpreters but it works like a compiler and generates fast analysis code?}
Similar to the concrete interpreter, for an abstract interpreter
$\widehat{\texttt{eval}}$ that takes a program, an abstract environment and
returns abstract values, shown as the following:

\begin{lstlisting}
def $\widehat{\texttt{eval}}$(e: Expr)($\rho$: $\widehat{\texttt{Env}}$): $\widehat{\texttt{Value}}$ = ...
\end{lstlisting}

We would like to have a specialized abstract interpreter
$\widehat{\texttt{eval}}_{\texttt{e0}}$ given a program $e_0$ statically, but
leave the abstract environment $\rho$ dynamic.

\begin{figure}[h]
  \begin{tikzpicture}
  \matrix (m) [matrix of math nodes,row sep=3em,column sep=4em,minimum width=2em]
  {
    \begin{smallmatrix} \text{unstaged} \\ \text{abstract} \\ \text{interpreters} \end{smallmatrix} & \begin{smallmatrix} \text{staged} \\ \text{abstract} \\ \text{interpreters} \end{smallmatrix} \\
      \begin{smallmatrix} \text{unstaged} \\ \text{concrete} \\ \text{interpreters} \end{smallmatrix} & \begin{smallmatrix} \text{staged} \\ \text{concrete} \\ \text{interpreters} \end{smallmatrix} \\ };
  \path[-stealth]
    (m-1-1) edge node [above, font=\scriptsize] {$\updownarrows$} (m-1-2)
    (m-2-1) edge node [left, font=\scriptsize] {$\alpha$} (m-1-1)
    (m-2-1) edge node [below, font=\scriptsize] {$\updownarrows$} (m-2-2)
    (m-2-2) edge node [right, font=\scriptsize] {$\alpha$} (m-1-2);
  \end{tikzpicture}
  \caption{The confluence of specialization and approximation.}
  \label{confluence}
\end{figure}

@@

Besides the intellectual merit of the confluence diagram, this paper contributes
to a practical problem in static analysis: \textbf{can we speed up a static analyzer
without any intrusive changes and soundness compromises?} Static analysis is
known as a trade-off game between precision and efficiency, we argue that
applying meta-programming (particularly, multi-stage programming) with abstract
interpreters is an effective approach to improve the latter one while fix the
former one. Specialization removes the overhead of abstractions introduced by
high-level programming, but not changes the meaning of the program -- in this
case, the abstract semantics implemented by the analyzer.
Indeed, specializing analysis with respect to a program by partial evaluation
is not new \cite{damian1999partial, amtoft1999partial,
  Boucher:1996:ACN:647473.727587, ashley:practical}, despite the fact that these
work are ad-hoc on a particular analysis or requiring significant changes to
the analysis. For example, abstract compilation \cite{Boucher:1996:ACN:647473.727587}
requires the whole analyzer to be rewritten as closure generation from. In this
paper, we show a systematic approach to meta-programming achieves the
performance goal with less intrusive changes; in fact, a case study
(Section~\ref{cs_ac}) that compared with abstract compilation shows, by
utilizing type-directed binding-time annotations, the analyzer program does not
need any changes.

\todo{modular analysis}

\iffalse
Likewise, specializing static analyses by partial evaluation emerged in late 90s
, and indeed it is able to effectively remove the interpretive
overhead of repeatedly traversing the abstract syntax tree. However, these
previous works focused mostly on one particular analysis, or required to
completely rewrite the analyzer. Hence, it is worth to investigate the idea from
a modern perspective based on generative programming, especially for a general
abstract interpreter that models direct-style $\lambda$-calculus, imperative
features such as mutation, and different abstract domains. One technical
challenge is the binding-time engineering when non-determinism, fixed-point
iterations, different abstract domains and staging are introduced at the same
time. In this paper, we present an end-to-end design and implementation of
staging an abstract interpreter; that means not only the interpreter that
traverses the abstract syntax tree, but also the data structure of abstract
domains, abstract environment and heap, and the fixed-point iteration are
staged.

@@

When staging a concrete interpreter, the programmers need to distinguish static
and dynamic values --- the given program to be executed by the interpreter is
classified as static because it is known at compile-time, and the inputs to that
program are dynamic. However, when staging an abstract interpreter, this
distinction does not exist anymore. Because the abstract interpreter
instantiates all the inputs as some form of abstract values, which are usually
top elements in their abstract domains and are also statically known. Then what
is the point of staging if there is no such distinction? A surprising by-product
of thinking about this question is to realize that we can apply the staged
abstract interpreter on \textit{open} programs, and the free variables
representing other parts of the program (e.g., libraries) are dynamic inputs,
therefore we obtain a modular analysis through staging, mechanically.
\note{TR: I don't understand this. The program structure is static, the abstract values
are still dynamic, no? They change in every iteration of the fixpoint algorithm}
\note{TR: why is it a big deal. On Closed programs we obtain a constant factor,
but modular analysis has different asymptotics}
\fi

### 2019-02-19 paper/70_related.tex
intro

>>>>

Multi-stage programming has been widely used to improve the performance in many
domains, such as optimizing compilers or domain-specific languages
\cite{DBLP:conf/pldi/RompfSBLCO14, DBLP:conf/snapl/RompfBLSJAOSKDK15,
DBLP:journals/tecs/SujeethBLRCOO14, DBLP:conf/gpce/SujeethGBLROO13,
DBLP:journals/jfp/CaretteKS09}, numerical computation \cite{PGL-038,
DBLP:conf/pepm/AktemurKKS13}, generic programming
\cite{DBLP:journals/pacmpl/Yallop17, Ofenbeck:2017:SGP:3136040.3136060}, data
processing \cite{DBLP:conf/oopsla/JonnalageddaCSRO14,
DBLP:conf/popl/KiselyovBPS17}, query compilation in databases
\cite{DBLP:conf/osdi/EssertelTDBOR18, DBLP:conf/sigmod/TahboubER18}, etc.
